---
title: "Advanced ggplot2"
subtitle: "Scientific workflows: Tools and Tips 🛠️"
author: "Dr. Selina Baldauf"
date: "2026-05-21"
format:
  revealjs:
    footer: "Selina Baldauf // Advanced ggplot2"
    highlight-style: printing
    mainfont: Cabinet Grotesk
    slide-number: true
    show-slide-number: all
    incremental: true
    self-contained: true
    code-line-numbers: false
    theme: slides.scss
    auto-stretch: false
editor: source
execute:
  echo: false
  eval: true
  cache: true
  message: false
from: markdown+emoji
---

```{r}
#| label: setup
#| include: false
#| cache: false

library(ggplot2)
library(dplyr)
library(gapminder)
library(patchwork)
library(scales)
library(gghighlight)
library(ggrepel)
library(ggtext)

gap_2007 <- gapminder |> filter(year == 2007)

# The custom theme we'll build in module 1
theme_workshop <- function() {
  theme_minimal(base_size = 12) %+replace%
    theme(
      plot.title = element_text(
        size = 14, hjust = 0,
        margin = margin(b = 10)
      ),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90"),
      legend.position = "bottom",
      plot.background = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA)
    )
}

p_default <- ggplot(
  gap_2007,
  aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)
) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  labs(
    title = "Life expectancy vs GDP per capita (2007)",
    x = "GDP per capita (USD)",
    y = "Life expectancy (years)",
    color = "Continent",
    size = "Population"
  )

# data and palette for module 2
gap_continent <- gapminder |>
  group_by(year, continent) |>
  summarise(mean_lifeExp = mean(lifeExp), .groups = "drop")

okabe_ito_continent <- c(
  Africa = "#E69F00",
  Americas = "#56B4E9",
  Asia = "#009E73",
  Europe = "#0072B2",
  Oceania = "#D55E00"
)

p_continent <- ggplot(
  gap_continent,
  aes(x = year, y = mean_lifeExp, color = continent)
) +
  geom_line(linewidth = 1) +
  labs(
    title = "Life expectancy over time by continent",
    x = NULL,
    y = "Life expectancy (years)",
    color = NULL
  )

# data and helpers for module 3
gap_asia <- gapminder |> filter(continent == "Asia")

highlight_colors_asia <- c(China = "#E69F00", India = "#0072B2")

endpoints_asia <- gap_asia |>
  filter(year == max(year), country %in% c("China", "India"))

# named plots for module 4
p_scatter <- ggplot(
  gap_2007,
  aes(x = gdpPercap, y = lifeExp, color = continent)
) +
  geom_point(alpha = 0.7, size = 3) +
  scale_x_log10(labels = label_dollar()) +
  labs(title = "2007 snapshot", x = "GDP per capita", y = "Life expectancy (years)")

p_lines <- ggplot(
  gap_continent,
  aes(x = year, y = mean_lifeExp, color = continent)
) +
  geom_line(linewidth = 1) +
  labs(title = "Over time", x = NULL, y = "Life expectancy (years)")

p_box <- ggplot(
  gap_2007,
  aes(x = continent, y = lifeExp, color = continent)
) +
  geom_boxplot(fill = "white") +
  labs(title = "By continent (2007)", x = NULL, y = "Life expectancy (years)")
```

## What is this lecture series?

### Scientific workflows: Tools and Tips :hammer_and_wrench:

::: nonincremental

:date: Every 3rd Thursday :clock4: 4-5 p.m. :round_pushpin: Webex

- One topic from the world of scientific workflows
- Material provided [online](https://selinazitrone.github.io/tools_and_tips/)
- If you don't want to miss a lecture
  - [Subscribe to the mailing list](https://lists.fu-berlin.de/listinfo/toolsAndTips)
- For credit points: Send me a short message (Email or Webex)

:::

## Why this workshop?

Default `ggplot2` plots look fine for exploration

. . .

...but they're rarely publication-ready out of the box:

- Default themes carry decoration you didn't ask for
- Default colors aren't colorblind-safe
- Text sizes don't match any output medium
- Single plots — papers want A/B/C panels
- Exports come out blurry or the wrong size

. . .

Today: a tour of the parts of the `ggplot2` ecosystem that fix these

## Plan for today

::: nonincremental

1. **Themes** — write your own and reuse it
2. **Color** — colorblind-safe palettes that stick to your groups
3. **Directing attention** — `ggtext`, `ggrepel`, `gghighlight`
4. **Multi-panel layouts** — A/B/C figures with `patchwork`
5. **Export** — get fonts and dimensions right for journals

Plus a short tour of useful extension packages at the end.

:::

## Follow along

::: nonincremental

Everything we do today is in this repo:

[github.com/selinazitrone/advanced-ggplot-workshop](https://github.com/selinazitrone/advanced-ggplot-workshop)

1. Clone it (or download the zip)
2. Open `advanced-ggplot-workshop.Rproj`
3. From the R console: `source("install_packages.R")`

For each module: I demo from `R/`, you do a short exercise from `exercises/`.

:::

# Build a reusable theme {.inverse}

## The default theme carries baggage

```{r}
#| label: plot-default
#| fig-width: 9
#| fig-height: 5
p_default
```

## A custom theme, applied

```{r}
#| label: plot-themed
#| fig-width: 9
#| fig-height: 5
p_default + theme_workshop()
```

## Anatomy of `theme()`

`theme()` controls every non-data element of the plot.

Elements form an inheritance tree:

```
text  ──┬── axis.text  ──┬── axis.text.x
        │                └── axis.text.y
        ├── plot.title
        ├── plot.subtitle
        └── legend.text

line  ──┬── axis.line   ──┬── axis.line.x
        │                 └── axis.line.y
        ├── axis.ticks
        └── panel.grid   ─── panel.grid.major / minor

rect  ──┬── plot.background
        └── panel.background
```

Setting a parent affects all its children.

## Cheatsheet: every theme element on one plot

A beautiful reference by Henry Wang showing every named element:

::: nonincremental

[henrywang.nl/ggplot2-theme-elements-demonstration](https://henrywang.nl/ggplot2-theme-elements-demonstration/)

:::

. . .

Keep this open while you tweak — it saves a lot of `?theme` lookups.

## Two ways to modify a theme

**With `+` :** keep the base, tweak it

```r
p_default +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

. . .

**With `%+replace%` :** start fresh, you specify everything

```r
my_theme <- theme_minimal() %+replace%
  theme(panel.grid.minor = element_blank())
```

. . .

Use `%+replace%` inside theme **functions** — full control, no surprises.

## Make it a function

```r
theme_workshop <- function() {
  theme_minimal(base_size = 12) %+replace%
    theme(
      plot.title = element_text(size = 14, hjust = 0),
      axis.title = element_text(size = 12),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position = "bottom"
      # ...
    )
}

p + theme_workshop()
```

::: nonincremental

- A function composes cleanly with other layers
- Easy to share, version, edit in one place

:::

## Apply globally, then source from a file

Set it once at the top of a script:

```r
theme_set(theme_workshop())
```

. . .

Every subsequent plot uses it — no per-plot `+ theme_workshop()`.

. . .

In a real project, the theme lives in its own file:

```r
# at the top of every analysis script
source(here::here("R", "theme_lab.R"))
theme_set(theme_lab())
```

One edit to `R/theme_lab.R` updates every figure in your project.

## Your turn :writing_hand: (8 min)

::: nonincremental

Open `exercises/01_themes.R`

1. Create `R/theme_lab.R` with a function `theme_lab()`:
   - based on `theme_minimal()`
   - no minor gridlines
   - axis title size 12, plot title size 14
2. Source it and apply with `theme_set(theme_lab())`

**Stretch goals** (if you finish early):

- Custom plot background color
- Google font via `sysfonts` + `showtext`
- Make a second plot and confirm the theme applies automatically

:::

# Color with intent {.inverse}

## Default colors carry baggage too

```{r}
#| label: line-default
#| fig-width: 9
#| fig-height: 5
p_continent
```

## With a colorblind-safe palette

```{r}
#| label: line-okabe
#| fig-width: 9
#| fig-height: 5
p_continent +
  scale_color_manual(values = okabe_ito_continent)
```

## Pick intuitive colors when you can

If a meaning is already in your reader's head, use it:

::: nonincremental

- Water, sky → blue
- Blood, fire → red
- Vegetation → green
- Gold, sun → yellow

:::

. . .

If there's no natural mapping, pick from a safe palette and stay consistent across all figures in the paper.

## Three safe palette families

::: nonincremental

- **Okabe-Ito** — 8 hand-picked categorical colors, distinguishable for all common forms of colorblindness
- **viridis** — perceptually uniform, great for continuous or many groups (`scale_*_viridis_d/c()`)
- **ColorBrewer (CB-safe subset)** — qualitative, sequential, diverging palettes (`scale_*_brewer()`)

:::

. . .

Browse hundreds more with **`paletteer`** — a meta-package for ~2500 palettes from dozens of sources.

## Check it: `cvd_grid()`

```r
library(colorblindr)
cvd_grid(p)
```

Renders your plot under four simulations:

::: nonincremental

- Deuteranopia (red-green)
- Protanopia (red-green)
- Tritanopia (blue-yellow)
- Grayscale

:::

. . .

If you can still read the plot in all four, you're good.

## Named vectors: colors that stick

**Positional** (fragile):

```r
scale_color_manual(values = c("orange", "blue", "green", "navy", "red"))
```

If one continent drops out of your filtered data, every other continent shifts colors.

. . .

**Named** (defensive):

```r
scale_color_manual(values = c(
  Africa = "#E69F00",
  Americas = "#56B4E9",
  Asia = "#009E73",
  Europe = "#0072B2",
  Oceania = "#D55E00"
))
```

Asia is always green, no matter which subset you plot.

## Double-coding: color + shape

When you have ≤4 categories, encode the same group twice:

```r
aes(color = group, shape = group)
# or
aes(color = group, linetype = group)
```

. . .

```{r}
#| label: double-coding
#| fig-width: 9
#| fig-height: 4

gap_three <- gap_continent |>
  filter(continent %in% c("Africa", "Asia", "Europe"))

ggplot(
  gap_three,
  aes(x = year, y = mean_lifeExp, color = continent, linetype = continent)
) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = okabe_ito_continent) +
  labs(x = NULL, y = "Life expectancy (years)", color = NULL, linetype = NULL)
```

Useful for colorblind viewers and for B/W print.

## Your turn :writing_hand: (8 min)

::: nonincremental

Open `exercises/02_color.R`

1. Build a named Okabe-Ito vector for the 5 continents and apply with `scale_color_manual()`
2. Run `cvd_grid()` and check it survives all four simulations

**Stretch goals**:

- Try 2-3 other palettes from `paletteer` and check each with `cvd_grid()` — which survives best?
- Make a continuous version: color points by `gdpPercap` with `scale_color_viridis_c()`

:::

# Directing attention {.inverse}

## Too many lines, no story

```{r}
#| label: asia-default
#| fig-width: 9
#| fig-height: 5
ggplot(gap_asia, aes(x = year, y = gdpPercap, color = country)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "GDP per capita in Asia, 1952-2007", x = NULL, y = "GDP per capita")
```

## `gghighlight`: dim the rest in one line

```{r}
#| label: asia-highlight
#| fig-width: 9
#| fig-height: 4.5
ggplot(gap_asia, aes(x = year, y = gdpPercap, color = country)) +
  geom_line(linewidth = 1) +
  gghighlight(country %in% c("China", "India"), use_direct_label = FALSE) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "GDP per capita in Asia, 1952-2007", x = NULL, y = "GDP per capita")
```

```r
gghighlight(country %in% c("China", "India"))
```

## Control the highlighted colors

```{r}
#| label: asia-colored
#| fig-width: 9
#| fig-height: 4.5
ggplot(gap_asia, aes(x = year, y = gdpPercap, color = country)) +
  geom_line(linewidth = 1) +
  gghighlight(country %in% c("China", "India"), use_direct_label = FALSE) +
  scale_color_manual(values = highlight_colors_asia) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "GDP per capita in Asia, 1952-2007", x = NULL, y = "GDP per capita")
```

## `ggrepel`: labels at the endpoints

```{r}
#| label: asia-labeled
#| fig-width: 9
#| fig-height: 4.5
ggplot(gap_asia, aes(x = year, y = gdpPercap, color = country)) +
  geom_line(linewidth = 1) +
  gghighlight(country %in% c("China", "India"), use_direct_label = FALSE) +
  scale_color_manual(values = highlight_colors_asia) +
  scale_y_continuous(labels = label_dollar()) +
  geom_text_repel(
    data = endpoints_asia, aes(label = country),
    nudge_x = 3, hjust = "left", direction = "y", segment.color = NA
  ) +
  labs(title = "GDP per capita in Asia, 1952-2007", x = NULL, y = "GDP per capita") +
  theme(legend.position = "none")
```

## `ggtext`: colored words in the title

```{r}
#| label: asia-final
#| fig-width: 9
#| fig-height: 4.5
ggplot(gap_asia, aes(x = year, y = gdpPercap, color = country)) +
  geom_line(linewidth = 1) +
  gghighlight(country %in% c("China", "India"), use_direct_label = FALSE) +
  scale_color_manual(values = highlight_colors_asia) +
  scale_y_continuous(labels = label_dollar()) +
  geom_text_repel(
    data = endpoints_asia, aes(label = country),
    nudge_x = 3, hjust = "left", direction = "y", segment.color = NA
  ) +
  labs(
    title = "**<span style='color:#E69F00'>China</span>** and **<span style='color:#0072B2'>India</span>**: GDP per capita 1952-2007",
    x = NULL, y = "GDP per capita"
  ) +
  theme(legend.position = "none", plot.title = element_markdown())
```

## `annotate()`: one-off notes and arrows

For a labelled event on the plot:

```r
+ annotate("label", x = 1960, y = 55000,
           label = "China's reform period\nbegan in 1978",
           label.size = 0, fill = "white") +
  annotate("segment", x = 1972, xend = 1978, y = 50000, yend = 3000,
           arrow = arrow(length = unit(0.2, "cm")))
```

## A sprinkle: `scales` for axis labels

You already saw `scales::label_dollar()` on the y-axis. The whole family is useful:

```r
scale_y_continuous(labels = label_dollar())
scale_y_continuous(labels = label_comma())
scale_y_continuous(labels = label_percent())
scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()))
```

The last renders `1200000` as `"1.2M"`.

## Your turn :writing_hand: (10 min)

::: nonincremental

Open `exercises/03_attention.R`

1. Find the top 5 GDP-growth countries (1952 → 2007) and highlight them with `gghighlight`; dim the rest
2. Add `ggrepel` endpoint labels at year 2007
3. Drop the legend

**Stretch goals**:

- Color the 5 with a named Okabe-Ito vector
- `ggtext` title naming the 5 in their colors
- `annotate()` a context note

:::

## Coffee break :coffee: (5 min)

# Multi-panel layouts with patchwork {.inverse}

## Three plots, named as R objects

```r
p_scatter <- ggplot(...) + ...
p_lines   <- ggplot(...) + ...
p_box     <- ggplot(...) + ...
```

. . .

Building plots as named objects is the small habit that unlocks composition. You'll also stop re-running long pipelines just to tweak one layer.

## Side by side, stacked, nested

```r
p1 + p2          # side by side
p1 / p2          # stacked
p1 | p2          # alias for +
(p1 + p2) / p3   # nested
```

## Combine into A/B/C

```{r}
#| label: patch-basic
#| fig-width: 10
#| fig-height: 5.5
(p_scatter + p_lines) / p_box
```

## Collect legend and axis titles

```{r}
#| label: patch-collect
#| fig-width: 10
#| fig-height: 5.5
(p_scatter + p_lines) / p_box +
  plot_layout(guides = "collect", axis_titles = "collect") &
  theme(legend.position = "bottom")
```

## Tag panels automatically

```{r}
#| label: patch-tags
#| fig-width: 10
#| fig-height: 5.5
(p_scatter + p_lines) / p_box +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```

## `&` vs `+` — apply to all panels

`+` modifies only the last plot in the chain.

`&` applies to **every** plot:

```r
(p1 + p2) / p3 +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

## `inset_element()`

```{r}
#| label: patch-inset
#| fig-width: 9
#| fig-height: 5
p_scatter +
  inset_element(
    p_box + theme_void() + theme(legend.position = "none"),
    left = 0.55, bottom = 0.05, right = 0.98, top = 0.45
  )
```

## Your turn :writing_hand: (8 min)

::: nonincremental

Open `exercises/04_patchwork.R`

1. Combine the three provided plots into an A/B/C figure with panel tags, one collected legend, shared y-axis title where it makes sense

**Stretch goals**:

- Compare `(p_scatter + p_lines) / p_box` vs `p_scatter / (p_lines + p_box)` — which tells the story better?
- Set `widths` so `p_scatter` is twice as wide as `p_lines`
- Add an `inset_element()` with a small overview plot in the corner of one panel

:::

# Export for journals {.inverse}

## What you see is not what you get

Default `ggsave("plot.png", p)`:

- uses your IDE's plot pane dimensions
- uses the system antialiaser — fuzzy text on many setups

. . .

What you want:

```r
ggsave(
  "fig1.png", p,
  width = 89, height = 70,
  units = "mm", dpi = 300
)
```

## Real journal column widths

::: nonincremental

- PLOS-Biology single column ≈ **89 mm**
- PLOS-Biology double column ≈ **183 mm**
- Each journal publishes its specs — check the figure guidelines

:::

. . .

Specify dimensions in **mm** so they match what the journal expects.

## The text-size trap

A title at size 14 looks fine in your plot pane (~7 inches wide) but tiny at 89 mm.

. . .

Text scales with physical size, not pixel count. Iterate font sizes at the real export dimensions, not at your default pane size.

## `ggview`: preview at export size

```r
library(ggview)
ggview(p, width = 89, height = 70, units = "mm")
```

. . .

Wraps your plot and previews it at the requested dimensions in the plot pane. No save-and-reopen cycle.

## Why `ragg::agg_png` and not plain `png()`?

```r
ggsave("fig1.png", p, device = ragg::agg_png, ...)
```

::: nonincremental

- AGG renderer → better antialiasing, crisper text
- Handles unicode, emoji, ligatures cleanly via `systemfonts`
- Consistent across Windows / macOS / Linux

:::

. . .

For PDF: `device = cairo_pdf` — embeds fonts so reviewers see what you saw.

## Online vs print

::: nonincremental

- Design for **print** (mm + dpi)
- The PDF is the canonical format; most readers download it
- Online HTML readers rescale that same image
- For talks and social media, re-export at slide-deck dimensions with bigger text

:::

## Your turn :writing_hand: (5 min)

::: nonincremental

Open `exercises/05_export.R`

1. Export the provided plot with default `ggsave("default.png", p)`
2. Export it with `ragg::agg_png` at PLOS single column (89 mm × 70 mm, 300 dpi)
3. Open both files, zoom in, compare the text

**Stretch goals**:

- Export at double-column width (183 mm) — what changes about font sizes?
- Export as `cairo_pdf`; check fonts are embedded (Acrobat → Properties → Fonts)
- Use `ggview()` to iterate

:::

# Honorable mentions {.inverse}

## `ggdist`

Halfeye, raincloud, dotsinterval. The modern bar-plot replacement.

```{r}
#| label: hm-ggdist
#| fig-width: 8
#| fig-height: 3.5
ggplot(gap_2007, aes(x = continent, y = lifeExp)) +
  ggdist::stat_halfeye(adjust = 0.5, width = 0.5, justification = -0.2, .width = 0) +
  geom_boxplot(width = 0.12, outlier.shape = NA)
```

```r
ggdist::stat_halfeye(...) + geom_boxplot(...)
```

## `ggsignif` / `ggpubr`

Significance brackets and p-values on plots.

```{r}
#| label: hm-ggsignif
#| fig-width: 7
#| fig-height: 3.5
ggplot(
  gap_2007 |> filter(continent %in% c("Asia", "Europe", "Africa")),
  aes(x = continent, y = lifeExp)
) +
  geom_boxplot() +
  ggsignif::geom_signif(
    comparisons = list(c("Asia", "Europe"), c("Asia", "Africa")),
    step_increase = 0.1
  )
```

```r
ggsignif::geom_signif(comparisons = list(c("Asia", "Europe")))
```

*(Some statisticians dislike on-figure p-values — know the arguments.)*

## `ggridges`

Ridgeline plots for distributions across groups or time.

```{r}
#| label: hm-ggridges
#| fig-width: 8
#| fig-height: 4
ggplot(gapminder, aes(x = lifeExp, y = continent, fill = continent)) +
  ggridges::geom_density_ridges(alpha = 0.7) +
  theme(legend.position = "none")
```

```r
ggridges::geom_density_ridges()
```

## `ggh4x`

Nested facets and custom strip styling for factorial designs.

```{r}
#| label: hm-ggh4x
#| fig-width: 9
#| fig-height: 3.8
ggplot(
  gapminder |> filter(
    year %in% c(1957, 1987, 2007),
    continent %in% c("Asia", "Europe", "Africa")
  ),
  aes(x = gdpPercap, y = lifeExp)
) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  ggh4x::facet_nested(~ continent + year)
```

```r
ggh4x::facet_nested(~ continent + year)
```

## `ggExtra`

Marginal histograms / densities on a scatter.

```{r}
#| label: hm-ggmarginal
#| fig-width: 7
#| fig-height: 4.5
p_marg <- ggplot(
  gap_2007,
  aes(x = gdpPercap, y = lifeExp, color = continent)
) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  theme(legend.position = "none")

ggExtra::ggMarginal(p_marg, type = "histogram", groupColour = TRUE, groupFill = TRUE)
```

```r
ggExtra::ggMarginal(p, type = "histogram")
```

## `cowplot`

Patchwork's older sibling. Useful helpers like `plot_grid()`, `theme_cowplot()`, `draw_plot()`.

```{r}
#| label: hm-cowplot
#| fig-width: 9
#| fig-height: 3.5
cowplot::plot_grid(
  p_scatter + theme(legend.position = "none"),
  p_lines + theme(legend.position = "none"),
  labels = c("A", "B"),
  nrow = 1
)
```

```r
cowplot::plot_grid(p1, p2, labels = c("A", "B"))
```

# Wrap up {.inverse}

## Recap

::: nonincremental

The inside-out arc:

1. **Theme** — strip default decoration, define your own
2. **Color** — encode meaning, stay accessible
3. **Attention** — direct the eye where it matters
4. **Layout** — compose multi-panel figures
5. **Export** — ship at the right size and resolution

:::

## Keep learning

::: nonincremental

- [Cédric Scherer's tutorials](https://www.cedricscherer.com/) — the modern ggplot2 reference
- [Albert Rapp's blog](https://albert-rapp.de/) — practical, polished examples
- [ggplot2 book (3rd ed)](https://ggplot2-book.org/) — the canonical reference
- [R Graphics Cookbook](https://r-graphics.org/) — recipe-style problems and answers

:::

## Next lecture

#### Topic t.b.a.

<br>

:date: 18th June :clock3: 4 - 5 p.m. :round_pushpin: Webex

:bell: [Subscribe to the mailing list](https://lists.fu-berlin.de/listinfo/toolsAndTips)

:e-mail: For topic suggestions or feedback [send me an email](mailto:selina.baldauf@fu-berlin.de)

## The end :)

Questions?
