---
title: "Stop repeating yourself: functions and iteration in R"
subtitle: "Scientific workflows: Tools and Tips 🛠️"
date: "2026-06-18"
format:
  revealjs:
    footer: "Selina Baldauf // Functions and iteration in R"
    highlight-style: printing
    mainfont: Cabinet Grotesk
    slide-number: true
    show-slide-number: all
    incremental: true
    self-contained: true
    code-line-numbers: false
    theme: slides.scss
    auto-stretch: false
editor: source
execute:
  echo: true
  eval: false
from: markdown+emoji
---

## What is this lecture series?

### Scientific workflows: Tools and Tips :hammer_and_wrench:

::: nonincremental

:date: Every 3rd Thursday :clock4: 4-5 p.m. :round_pushpin: Webex

- One topic from the world of scientific workflows
- Material provided [online](https://selinazitrone.github.io/tools_and_tips/)
- If you don't want to miss a lecture
  - [Subscribe to the mailing list](https://lists.fu-berlin.de/listinfo/toolsAndTips)
- For credit points: Send me a short message (Email or Webex)

:::

## The problem: copy-paste code

:::{.columns}

:::{.column width="45%"}

- Copy paste code violates the **DRY** (do not repeat yourself) principle

- **Fragile** and hard to **maintain**
  - Copy-paste errors
  - Change logic in multiple places

:::{.fragment}

:::{.nonincremental}

**Solution**: Write a **function** to capture the logic, then **iterate** over different inputs

:::

:::

:::

:::{.column width="55%"}

```{r}
#| label: messy_script
library(tidyverse)
library(palmerpenguins)

# Adelie species
adelie <- filter(penguins, species == "Adelie")
ggplot(adelie, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

# Chinstrap species
chinstrap <- filter(penguins, species == "Chinstrap")
ggplot(chinstrap, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

# Gentoo species
gentoo <- filter(penguins, species == "Chinstrap")
ggplot(gentoo, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

:::

:::

## Today: Functions and iterations

:::{.nonincremental}

- How to write your own **functions**
- How to **iterate** using the `purrr` package from the `tidyverse`

:::

. . .

#### How it works

:::{.nonincremental}

- There's a **GitHub repo** you download and open in your IDE
- I demonstrate each topic: **follow along or just watch**
- Each module has a short **exercise** to try it yourself
- **Questions** anytime: in the chat or just unmute

:::

## Get set up

:::{.nonincremental}

1. Click the green **Code** button → **Download ZIP**
2. Unzip it somewhere you can find it
3. Double-click the `.Rproj` file → opens in RStudio
   - *(Positron/VS code users: open the folder)*
4. Run `install_packages.R` to install today's packages

```r
install.packages(c("tidyverse", "palmerpenguins", "here"))
```

Repo: [https://github.com/selinaZitrone/functions-iteration-workshop](https://github.com/selinaZitrone/functions-iteration-workshop)

:::

# Writing your own functions in R {.inverse}

## Anatomy of a function

A function has

- a **name**: Choose something descriptive that starts with a verb
- **arguments** (its inputs)
- a **body**: All the code that is run using the input
- a **return value**: the output of the function (the last line of the body)

:::{.columns}

:::{.column}

```r
name <- function(arguments) {
  body
}
```
:::

:::{.column}

:::{.fragment}

```{r}
#| label: function_example
#| eval: true
add_numbers <- function(x, y) {
  x + y
}

add_numbers(1, 1)
add_numbers(x = 1, y = 1)
```

:::
:::
:::

## Turn code into functions

- What **changes** becomes an **argument** (the data)
- What **stays the same** becomes the **body** (the ggplot)

::: {.columns}

::: {.column}

```{r}
adelie <- filter(penguins, species == "Adelie")
ggplot(adelie, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

:::

::: {.column}

```{r}
# generic penguin function
plot_penguins <- function(data) {
  ggplot(data, aes(flipper_length_mm, body_mass_g)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_minimal()
}
# apply the function to the Adelie data
adelie <- filter(penguins, species == "Adelie")
plot_penguins(adelie)
```


:::

:::

## Default arguments

- Arguments with default values don't have to be specified when calling the function
- They can be used for things that usually stay the same, but sometimes are changed
- Default arguments are defined in the function body with `arg = default`

```{r}
# Define a default color for penguin data points
plot_penguins <- function(data, point_color = "blue") {
  ggplot(data, aes(flipper_length_mm, body_mass_g)) +
    geom_point(color = point_color) +
    geom_smooth(method = "lm") +
    theme_minimal()
}
# Call the function with the default blue color
plot_penguins(adelie)
# Call the function with a custom color
plot_penguins(adelie, point_color = "red")
```


## Live demo {.inverse}

Build `plot_penguins()`

`demo/01_functions.R`

## Your turn — Exercise 1

Turn a different plot block into a function.

:::{.nonincremental}

- You're given a working boxplot — **wrap it** in a function `plot_mass_by_sex(data)`
- Call it for a different species
- **Going further:** add an argument with a default (`fill` or `title`)

:::

`exercises/01_functions_exercise.R`

# 2. Iterating with `map()` {.inverse}

## The classic tool: a for loop

"Do this for each element":

```r
roots <- c()
for (x in 1:3) {
  roots[x] <- sqrt(x)
}
roots
```

. . .

It works — but there's boilerplate: a container, an index, the filling-in. You'll see loops in the wild; today we use `map()`.

## `map()`: the one idea

```r
map(1:3, sqrt)        # a list, one result per element
```

. . .

`map(x, f)` calls `f` **once per element** of `x` and gives back **one result per element**.

. . .

`x` can be anything with elements — a vector, the **columns of a data frame**, a list of **file paths**. Only the *element* changes from example to example.

## The `map_*()` family

The suffix says what comes **back**:

::: nonincremental

- `map()` → list
- `map_dbl()` → double &nbsp; `map_chr()` → character
- `map_lgl()` → logical &nbsp; `map_int()` → integer

:::

. . .

All of these return something. `walk()` (Module 4) is the sibling for when you want a **side effect**, not a result.

## Anonymous functions: `\(x) ...`

Write a function on the spot when:

. . .

**There's no ready-made function**

```r
map_dbl(1:3, \(x) x^2)
```

. . .

**Your value isn't the first argument** — e.g. the species in `filter()`:

```r
map(species_names, \(s) filter(penguins, species == s))
```

## Live demo {.inverse}

Map over columns, write an anonymous function, and read every CSV in a folder.

`demo/02_map.R`

## Your turn — Exercise 2

Two ways to use `map()`:

:::{.nonincremental}

- **Task A** — map a ready-made function (no `\()`): distinct values per column
- **Task B** — an anonymous function: one data frame per **island**

:::

`exercises/02_map_exercise.R`

# 3. `map()` with your own functions {.inverse}

## Put the two skills together

We can plot one species:

```r
plot_penguins(filter(penguins, species == "Adelie"))
```

. . .

We want this for **every** species — that's `map`. But `map` needs a **list**: one data frame per species.

## `split()` into a named list

```r
by_species <- split(penguins, penguins$species)

names(by_species)        # "Adelie" "Chinstrap" "Gentoo"
by_species[["Adelie"]]   # each element is a data frame
```

. . .

A **named list of data frames** — exactly what `map` walks over.

## The payoff

```r
plots <- map(by_species, plot_penguins)
plots[["Adelie"]]
```

. . .

The messy three-block script is now **two lines**. Fix `plot_penguins` once, every plot updates — no copy-paste bug.

## The same shape, anything per group

```r
penguins |>
  split(penguins$species) |>
  map(\(df) lm(body_mass_g ~ flipper_length_mm, data = df)) |>
  map(summary) |>
  map_dbl(\(x) x$r.squared)
```

`split → map → map_dbl` works whether each element becomes a plot, a **model**, or a number. *(Bonus: `bonus/model_per_group.R`.)*

## Live demo {.inverse}

`demo/03_map_functions.R`

## Your turn — Exercise 3

One plot per **island** instead of per species.

:::{.nonincremental}

- The split is done for you — you write the `map`
- **Going further:** pull one number per island with `map_dbl()`, or map a function you wrote yourself

:::

`exercises/03_map_functions_exercise.R`

# 4. `walk()`: saving results {.inverse}

## `map` vs `walk`

- `map()` → you want the **results** back
- `walk()` → you want a **side effect** (print, save a file); the return value isn't the point

. . .

Same iteration idea, two purposes.

## Show them all, then save them all

```r
walk(plots, print)              # draw each plot
```

. . .

```r
save_plot <- function(plot, name) {
  ggsave(here("plots", paste0(name, ".png")), plot, width = 6, height = 4)
}

walk2(plots, names(plots), save_plot)   # plot + filename, in parallel
```

The names from `split()` become the filenames.

## Live demo {.inverse}

`demo/04_walk.R`

# Wrap-up {.inverse}

## The messy script, redeemed

```r
penguins_all <- list.files(here("data"), pattern = "\\.csv$", full.names = TRUE) |>
  map(read_csv) |>      # iterate over files  -> one data frame
  list_rbind()

by_species <- split(penguins_all, penguins_all$species)
plots <- map(by_species, plot_penguins)   # iterate over groups -> one plot each
walk2(plots, names(plots), save_plot)      # iterate to save     -> one file each
```

The whole analysis, one readable chain. Add a fourth species or a tenth file — nothing changes.

## Two skills, one shape

::: nonincremental

- **Functions:** write the logic once
- **Iteration:** run it over many inputs
  - `map()` for results, `walk()` for side effects

:::

. . .

A list → `map` your function → a result per element. That's it.

## Where next

::: nonincremental

- **More inputs at once:** `map2()`, `pmap()`
- **When some iterations fail:** `safely()`, `possibly()`
- **Going parallel:** purrr's `in_parallel()` (≥ 1.1.0) or the `furrr` package
- **Bare column names in your functions:** tidy-eval `{{ }}`

:::

## Thank you! {.inverse}

Write the thing once, then map it.

Questions?
