---
title: "Stop repeating yourself: functions and iteration in R"
subtitle: "Scientific workflows: Tools and Tips 🛠️"
date: "2026-06-18"
format:
  revealjs:
    footer: "Selina Baldauf // Functions and iteration in R"
    highlight-style: printing
    mainfont: Cabinet Grotesk
    slide-number: true
    show-slide-number: all
    incremental: true
    self-contained: true
    code-line-numbers: false
    theme: slides.scss
    auto-stretch: false
editor: source
execute:
  echo: true
  eval: false
from: markdown+emoji
---

```{r}
#| label: setup
#| include: false
#| eval: true
library(tidyverse)
library(palmerpenguins)
```

## What is this lecture series?

### Scientific workflows: Tools and Tips :hammer_and_wrench:

::: nonincremental

:date: Every 3rd Thursday :clock4: 4-5 p.m. :round_pushpin: Webex

- One topic from the world of scientific workflows
- Material provided [online](https://selinazitrone.github.io/tools_and_tips/)
- If you don't want to miss a lecture
  - [Subscribe to the mailing list](https://lists.fu-berlin.de/listinfo/toolsAndTips)
- For credit points: Send me a short message (Email or Webex)

:::

## The problem: copy-paste code

:::{.columns}

:::{.column width="45%"}

- Copy paste code violates the **DRY** (do not repeat yourself) principle

- Fragile and hard to maintain
  - Copy-paste errors
  - Change logic in multiple places

:::{.fragment}

:::{.nonincremental}

**Solution**: Write a **function** to capture the logic, then **iterate** over different inputs

:::

:::

:::

:::{.column width="55%"}

```{r}
#| label: messy_script
library(tidyverse)
library(palmerpenguins)

# Adelie species
adelie <- filter(penguins, species == "Adelie")
ggplot(adelie, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

# Chinstrap species
chinstrap <- filter(penguins, species == "Chinstrap")
ggplot(chinstrap, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

# Gentoo species
gentoo <- filter(penguins, species == "Chinstrap")
ggplot(gentoo, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

:::

:::

## Today: Functions and iterations

:::{.nonincremental}

- How to write your own **functions**
- How to **iterate** using the `purrr` package from the `tidyverse`

:::

. . .

#### How it works

:::{.nonincremental}

- There's a **GitHub repo** you download and open in your IDE
- I demonstrate each topic: **follow along or just watch**
- Each module has a short **exercise** to try it yourself
- **Questions** anytime: in the chat or just unmute

:::

## Get set up

:::{.nonincremental}

1. Click the green **Code** button → **Download ZIP**
2. Unzip it somewhere you can find it
3. Double-click the `.Rproj` file → opens in RStudio
   - *(Positron/VS code users: open the folder)*
4. Run `install_packages.R` to install today's packages

```r
install.packages(c("tidyverse", "palmerpenguins", "here"))
```

Repo: [https://github.com/selinaZitrone/functions-iteration-workshop](https://github.com/selinaZitrone/functions-iteration-workshop)

:::

# Writing your own functions in R {.inverse}

## Anatomy of a function

A function has

- a **name**: Choose something descriptive that starts with a verb
- **arguments** (its inputs)
- a **body**: All the code that is run using the input
- a **return value**: the output of the function (the last line of the body)

:::{.columns}

:::{.column}

:::{.fragment}

```r
name <- function(arguments) {
  body
}
```
:::
:::

:::{.column}

:::{.fragment}

```{r}
#| label: function_example
#| eval: true
add_numbers <- function(x, y) {
  x + y
}

add_numbers(1, 1)
add_numbers(x = 1, y = 1)
```

:::
:::
:::

## Turn code into functions

:::{.nonincremental}

- What **changes** becomes an **argument** (the data)
- What **stays the same** becomes the **body** (the ggplot)

:::

```{r}
adelie <- filter(penguins, species == "Adelie")
ggplot(adelie, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

. . .

The function version:

```{r}
# generic penguin function
plot_penguins <- function(data) {
  ggplot(data, aes(flipper_length_mm, body_mass_g)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_minimal()
}
# apply the function to the Adelie data
adelie <- filter(penguins, species == "Adelie")
plot_penguins(adelie)
```

## Default arguments

- Arguments with default values don't have to be specified when calling the function
- They can be used for things that usually stay the same, but sometimes are changed
- Default arguments are defined in the function body with `arg = default`

. . .

```{r}
# Define a default color for penguin data points
plot_penguins <- function(data, point_color = "blue") {
  ggplot(data, aes(flipper_length_mm, body_mass_g)) +
    geom_point(color = point_color) +
    geom_smooth(method = "lm") +
    theme_minimal()
}
# Call the function with the default blue color
plot_penguins(adelie)
# Call the function with a custom color
plot_penguins(adelie, point_color = "red")
```


## Live demo {.inverse}

Build `plot_penguins()`

`demo/01_functions.R`

## Your turn {.inverse}

> :pencil2: **Exercise 1 (~7 min):**
>
> Open `exercises/01_functions_exercise.R`

# Iteration with the `purrr` package {.inverse}

## Base R iteration: for-loops and `apply()` 

Run the same function for multiple inputs, e.g.

```{r}
sqrt(1)
sqrt(2)
sqrt(3)
```

. . .

Base R iteration options:

```{r}
# for loop
for (x in 1:3) {
  sqrt(x)
}

# Functions from the apply family
lapply(1:3, sqrt) # a list
sapply(1:3, sqrt) # a vector
```

## The `purrr` package

- Part of the `tidyverse` packages
- Provides functions for functional programming and iteration
- Why use it?
  - More consistent than base R `apply` functions
  - More flexible than base R iteration
  - Integrates well in a tidyverse workflow

## The main function: `map`
`map(x, f)` calls `f` **once per element** of `x` and gives back **one result per element**.

```{r}
# x: a list or vector to iterate over
# f: a function to apply to each element of x
map(x, f)
```
<br>
. . .

```{r}
#| label: map_example
#| eval: true
# The map function always returns its result as a list
library(purrr)
map(1:3, sqrt)
```

## More complex situations

What if it's not so simple, e.g. my iterator is not the first argument?

```{r}
# round the number pi to a different number of digits
round(pi, digits = 1)
round(pi, digits = 2)
round(pi, digits = 3)
```

Just using `map` with `round` does not work -> Where does `pi` go?

```{r}
# iterator is passed to the first argument not as `digits`
map(1:3, round)
```

## Solution: Anonymous function

Solution: Write an anonymous function on the spot:

```{r}
map(1:3, \(x) round(pi, digits = x))
# same as
map(1:3, function(x) round(pi, digits = x))
```

:::{.nonincremental}

- `\(x)` syntax is a shortcut for `function(x)`
- `x` is the argument that takes the value of each element of the iterator

:::

## The `map_*()` variants

The suffix says what is the output of the function:

::: nonincremental

- `map()` returns a list
- `map_dbl()` returns a numeric vector
- `map_chr()` returns a character vector
- `map_lgl()` returns a logical vector 
- `map_int()` returns an integer vector
:::

For example:

```{r}
# return result as list (always works)
map(1:3, sqrt)
# return results as double vector (only works if results are single numbers)
map_dbl(1:3, sqrt)
```

## Live demo {.inverse}

Map over columns, write an anonymous function, and read every CSV in a folder.

`demo/02_map.R`

## Your turn {.inverse}

> :pencil2: **Exercise 1 (~7 min):**
>
> Open `exercises/02_map_exercise.R`

# `map()` with your own functions {.inverse}

## Put the two skills together

We can plot one species with our function from earlier:

```r
plot_penguins(filter(penguins, species == "Adelie"))
```

But how do we do this for all three species? -> Map needs a list of data frames, one per species.

## `split()` into a named list

The base R function `split()` does exactly that:

```{r}
#| eval: true
# split the penguins data into a named list of data frames, one per species
by_species <- split(penguins, penguins$species)

names(by_species)        # "Adelie" "Chinstrap" "Gentoo"
by_species[["Adelie"]]   # each element is a data frame
```


## `map()` over the list of data frames

```{r}
plots <- map(by_species, plot_penguins)
# Look at one individual plot
plots[["Adelie"]]
```

. . .

`map` works very well with the pipe operator, so we can do everything in one chain:

```{r}
plots <- penguins |>
  split(penguins$species) |>
  map(plot_penguins)
```

## The same structure, anything per group

Run a linear model for each species and pull out the R² values:

```r
penguins |>
  split(penguins$species) |>
  map(\(df) lm(body_mass_g ~ flipper_length_mm, data = df)) |>
  map(summary) |>
  map_dbl(\(x) x$r.squared)
```

## Live demo {.inverse}

`demo/03_map_functions.R`

## Your turn {.inverse}

> :pencil2: **Exercise 1 (~7 min):**
>
> Open `exercises/03_map_functions_exercise.R`

# Saving results with `walk()` {.inverse}

## `map` vs `walk`

Both functions work in the same way, but they have different purposes:

- `map()` → you want the **results** back
- `walk()` → you want a **side effect** (print a plot, save a file)

## Example: print all the plots

```{r}
# Create all plots and save them in a list -> results we want back
plots <- penguins |>
  split(penguins$species) |>
  map(plot_penguins)

# Take the list of plots and print them -> side effect we want (the printing)
walk(plots, print)              # draw each plot
```

## Example: save all the plots

We can also `walk()` over two inputs at the same time with `walk2()`: the list of plots (x) and their names (y).

```{r}
# Walk 2 iterates over 2 inputs (x, y) in parallel
walk2(plots, names(plots), function(x, y) {ggsave(
  here("plots", paste0(y, ".png")), x
)})
```

## Live demo {.inverse}

`demo/04_walk.R`

# Wrap-up {.inverse}

## Function

- Functions are reusable pieces of code that take inputs and return outputs
- Default arguments are defined with `arg = value` in the function body

For example

```r
# y is optional and defaults to 1
add_numbers <- function(x, y = 1) {
  x + y
}
```

## Iteration with the `purrr` package

`map(x, f)` applies function `f` to each element of `x` and returns a list

```r
map(1:3, sqrt) # returns a list of square roots
map_dbl(1:3, sqrt) # returns a numeric vector of square roots
```
<br>
You can use `map` with your own functions, and you can write anonymous functions on the spot if needed.

```r
map(1:3, \(x) round(pi, digits = x))
```
<br>
The `walk()` function is the same as map but for side effects (printing, saving files) instead of returning results.

```r
walk(plots, print) # prints each plot
```

## Where next

::: nonincremental

- More inputs at once: `map2()`, `pmap()`
- When some iterations fail: `safely()`, `possibly()`
- Going parallel: purrr's `in_parallel()` or the `furrr` package
:::

Resources: [purrr cheatsheet](https://opensource.posit.co/resources/cheatsheets/purrr/) and [purrr documentation](https://purrr.tidyverse.org/)

## Next lecture/workshop

#### Topic: Code sharing and reproducible repositories

<br>

:date: 16.07.2026 :clock4: 4-5 p.m. :round_pushpin: Webex

:bell: [Subscribe to the mailing list](https://lists.fu-berlin.de/listinfo/toolsAndTips)

:e-mail: For topic suggestions and/or feedback [send me an email](mailto:selina.baldauf@fu-berlin.de)

# The end :) {background-color="#2c2c2c"}

Questions?
