---
title: "Publication-ready ggplot2"
subtitle: "Scientific workflows: Tools and Tips 🛠️"
date: "2026-05-21"
format:
  revealjs:
    footer: "Selina Baldauf // Publication-ready ggplot2"
    highlight-style: printing
    mainfont: Cabinet Grotesk
    slide-number: true
    show-slide-number: all
    incremental: true
    self-contained: true
    code-line-numbers: false
    theme: slides.scss
    auto-stretch: false
editor: source
execute:
  echo: true
  eval: true
  warning: false
  message: false
from: markdown+emoji
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(gapminder)
library(scales)
library(patchwork)

# "Before" plots should show the ggplot2 default look
theme_set(theme_grey())

gap_2007 <- gapminder |> filter(year == 2007)

# The theme we build together in Module 1 (lives in solutions/theme.R)
theme_workshop <- function(base_size = 16, ink = "grey20", paper = "white") {
  theme_light(base_size = base_size, ink = ink, paper = paper) %+replace%
    theme(
      legend.text = element_text(size = rel(0.85)),
      panel.grid.minor = element_blank()
    )
}

# The recurring bubble chart (default colors, no theme layer)
p_bubble <- ggplot(
  gap_2007,
  aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)
) +
  geom_point() +
  scale_x_log10(labels = label_dollar(accuracy = 1)) +
  scale_size(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    x = "GDP per capita (USD, log scale)",
    y = "Life expectancy (years)",
    color = "Continent",
    size = "Population"
  )
```

## What is this lecture series?

### Scientific workflows: Tools and Tips :hammer_and_wrench:

::: nonincremental

:date: Every 3rd Thursday :clock4: 4-5 p.m. :round_pushpin: Webex

- One topic from the world of scientific workflows
- Material provided [online](https://selinazitrone.github.io/tools_and_tips/)
- If you don't want to miss a lecture
  - [Subscribe to the mailing list](https://lists.fu-berlin.de/listinfo/toolsAndTips)
- For credit points: Send me a short message (Email or Webex)

:::

## From default to publication-ready

- Making a first plot is easy, but making it publication-ready takes more work, choices and iterations

. . .

:::{.nonincremental}

So many things you can tune:

- The plot itself: which geoms to pick
- Themes, colours, fonts
- How to highlight your results
- Creating multi-panel plots
- Exporting at a readable size

:::

. . .

Different outputs (poster/talk/journal figure) require different tweaks

## Today

We'll focus on:

1. **Custom themes** for a consistent look
2. **Colour** choice
3. **Multi-panel layouts**: how to combine plots
4. **Exporting** plots at the right size and resolution


::: {.callout-note}

This is not an exhaustive list, there is so much more ggplot extension packages and tricks out there.

:::


## How it works

:::{.nonincremental}

- There's a **GitHub repo** you download and open in your IDE (I'll use Positron)
- I demonstrate each topic: **follow along or just watch**
- Each module has a short **exercise** to play with the concept
- **Questions** anytime: in the chat or just unmute

:::

## Get set up

:::{.nonincremental}

1. Click the green **Code** button → **Download ZIP**
2. Unzip it somewhere you can find it
3. Double-click the `.Rproj` file → opens in RStudio
   - *(Positron/VS code users: open the folder)*
4. Run `install_packages.R` to install today's packages

Repo: https://github.com/selinaZitrone/advanced-ggplot-workshop

:::

# 1. Custom themes {.inverse}

## The default isn't publication-ready

```{r}
#| echo: false
p_bubble
```

## Built-in themes: a quick win

Add a `theme_*()` layer and the whole look changes.

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5.5
mini <- function(th, name) {
  p_bubble +
    th +
    labs(title = name, x = NULL, y = NULL) +
    theme(legend.position = "none", plot.title = element_text(size = 13))
}
wrap_plots(
  list(
    mini(theme_grey(), "theme_grey()"),
    mini(theme_bw(), "theme_bw()"),
    mini(theme_light(), "theme_light()"),
    mini(theme_minimal(), "theme_minimal()"),
    mini(theme_classic(), "theme_classic()")
  ),
  nrow = 2
)
```

## Further customization with `theme()`

A theme controls everything that isn't data: Fonts, sizes, colours, grid lines, spacing, legends, ...

![ggplot2 theme-system cheatsheet (by [Clara Granell](https://github.com/claragranell/ggplot2/tree/main))](images/2026_05_21_publication-ready-ggplot/ggplot_theme_system_cheatsheet.png){fig-align="center" width="60%"}

## Inheritance of `theme` elements

```
text  ──┬── axis.text  ──┬── axis.text.x
        │                └── axis.text.y
        ├── plot.title
        ├── plot.subtitle
        └── legend.text

line  ──┬── axis.line   ──┬── axis.line.x
        │                 └── axis.line.y
        ├── axis.ticks
        └── panel.grid   ─── panel.grid.major / minor

rect  ──┬── plot.background
        └── panel.background
```

- Set a **parent** and it affects all its **children**
- E.g. Change `text` once and every piece of text updates

## How to change a theme

```{r}
p_bubble +
  theme_light(base_size = 18) +
  theme(
    legend.text = element_text(size = rel(0.85)),
    panel.grid.minor = element_blank()
  )
```

## Live demo

Open `demo/01_themes.R`

## Your turn {.inverse}

> :pencil2: **Exercise (~7 min):**
>
> Open `exercises/01_themes_exercise.R`

# 2. Colour with intent {.inverse}

```{r}
#| label: m2-setup
#| include: false
library(scico)

okabe <- palette.colors(palette = "Okabe-Ito")
continent_colors <- c(
  Africa = okabe[[2]],
  Americas = okabe[[3]],
  Asia = okabe[[4]],
  Europe = okabe[[6]],
  Oceania = okabe[[1]]
)

# strip of discrete colours
swatch_discrete <- function(cols) {
  df <- data.frame(x = seq_along(cols))
  ggplot(df, aes(x, 1, fill = factor(x))) +
    geom_tile(width = 0.95) +
    scale_fill_manual(values = unname(cols), guide = "none") +
    theme_void()
}

# smooth strip from many colours
swatch_ramp <- function(cols) {
  ggplot(data.frame(x = seq_along(cols)), aes(x, 1, fill = x)) +
    geom_raster() +
    scale_fill_gradientn(colours = cols, guide = "none") +
    coord_cartesian(expand = FALSE) +
    theme_void() +
    theme(plot.margin = margin(2, 4, 2, 4))
}

# titled ramps in a grid (ncol controls layout)
ramp_panel <- function(named_cols, ncol = 1) {
  strips <- lapply(names(named_cols), function(nm) {
    swatch_ramp(named_cols[[nm]]) +
      labs(title = nm) +
      theme(plot.title = element_text(size = 12, hjust = 0))
  })
  patchwork::wrap_plots(strips, ncol = ncol)
}
```

## Pick intuitive colours

Use colours your reader already associates with the thing.

![[Blogpost on colours by Lisa Charlotte Muth (Datawrapper)](https://www.datawrapper.de/blog/colors)](images/2026_05_21_publication-ready-ggplot/intuitive_colors.png)


## Two things that matter

For scientific figures:

:::{.nonincremental}

1. **Intuitive** colours that match what your reader expects
2. **Colourblind-safe** colours

:::

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 2.4
#| fig-align: center
set.seed(42)
cb_demo <- data.frame(
  x = runif(50, 0, 10),
  y = runif(50, 0, 5),
  group = rep(c("A", "B"), each = 25)
)
rg_cols <- c(A = "red", B = "forestgreen")
rg_deutan <- colorspace::deutan(rg_cols); names(rg_deutan) <- names(rg_cols)
rg_protan <- colorspace::protan(rg_cols); names(rg_protan) <- names(rg_cols)

p_normal <- ggplot(cb_demo, aes(x, y, color = group)) +
  geom_point(size = 6, alpha = 0.9) +
  scale_color_manual(values = rg_cols, guide = "none") +
  labs(title = "Normal vision") +
  theme_void() +
  theme(plot.title = element_text(size = 13, hjust = 0.5))

p_deutan <- p_normal +
  scale_color_manual(values = rg_deutan, guide = "none") +
  labs(title = "Deuteranopia")

p_protan <- p_normal +
  scale_color_manual(values = rg_protan, guide = "none") +
  labs(title = "Protanopia")

p_normal + p_deutan + p_protan + plot_layout(nrow = 1)
```

## Three kinds of palettes

:::: {.columns}

::: {.column width="33%"}
**Qualitative**

```{r}
#| echo: false
#| fig-width: 3.2
#| fig-height: 0.7
swatch_discrete(continent_colors)
```

Distinct categories
:::

::: {.column width="33%"}
**Sequential**

```{r}
#| echo: false
#| fig-width: 3.2
#| fig-height: 0.7
swatch_ramp(viridisLite::viridis(256))
```

```{r}
#| echo: false
#| fig-width: 3.2
#| fig-height: 0.7
swatch_discrete(viridisLite::viridis(7))
```

Ordered, low to high
:::

::: {.column width="33%"}
**Diverging**

```{r}
#| echo: false
#| fig-width: 3.2
#| fig-height: 0.7
swatch_ramp(scico::scico(256, palette = "vik"))
```

Above / below a midpoint
:::

::::

## Okabe-Ito (base R)

A colourblind-safe qualitative palette, built into base R.

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 1
swatch_discrete(okabe)
```

- 8 distinct hues plus black and grey, separable under all common CVD types
- access with `palette.colors(palette = "Okabe-Ito")`


[See here for details](https://clauswilke.com/dataviz/color-pitfalls.html#fig:palette-Okabe-Ito)

## viridis (built into ggplot2)

Perceptually uniform, colourblind- and greyscale-safe. Best for ordered or continuous data.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.4
ramp_panel(
  list(
    magma = viridisLite::viridis(256, option = "A"),
    inferno = viridisLite::viridis(256, option = "B"),
    plasma = viridisLite::viridis(256, option = "C"),
    viridis = viridisLite::viridis(256, option = "D"),
    cividis = viridisLite::viridis(256, option = "E"),
    rocket = viridisLite::viridis(256, option = "F"),
    mako = viridisLite::viridis(256, option = "G"),
    turbo = viridisLite::viridis(256, option = "H")
  ),
  ncol = 2
)
```

- access with `scale_colour_viridis_d()` / `_c()` 

[See here for details](https://ggplot2.tidyverse.org/reference/scale_viridis.html)

## scico R package

Perceptually uniform and colourblind-safe. Made for **scientific data** and designed to be fair, readable, and citable


```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.4
ramp_panel(
  list(
    batlow = scico::scico(256, palette = "batlow"),
    lajolla = scico::scico(256, palette = "lajolla"),
    bilbao = scico::scico(256, palette = "bilbao"),
    oslo = scico::scico(256, palette = "oslo"),
    vik = scico::scico(256, palette = "vik"),
    roma = scico::scico(256, palette = "roma"),
    berlin = scico::scico(256, palette = "berlin"),
    broc = scico::scico(256, palette = "broc")
  ),
  ncol = 2
)
```

- access with `scale_colour_scico_d()` / `_c()`

[scico R package](https://github.com/thomasp85/scico), the [colour maps by Fabio Crameri](https://www.fabiocrameri.ch/colourmaps/)

## Live demo

Open `demo/02_color.R`.

## Your turn {.inverse}

> :pencil2: **Exercise (~5 min):**
>
> Open `exercises/02_color_exercise.R`

# 3. Multi-panel layouts {.inverse}

```{r}
#| label: m4-setup
#| include: false
gap_continent <- gapminder |>
  group_by(year, continent) |>
  summarise(
    mean_lifeExp = mean(lifeExp),
    mean_gdpPercap = mean(gdpPercap),
    .groups = "drop"
  )

p_bubble_m4 <- ggplot(
  gap_2007,
  aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)
) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = label_dollar(accuracy = 1)) +
  scale_size(labels = label_number(scale_cut = cut_short_scale())) +
  scale_color_scico_d(palette = "batlow") +
  theme_workshop(base_size = 16) +
  labs(
    x = "GDP per capita (USD, log scale)",
    y = "Life expectancy",
    color = "Continent",
    size = "Population"
  )

p_life <- ggplot(gap_continent, aes(year, mean_lifeExp, color = continent)) +
  geom_line(linewidth = 1) +
  scale_color_scico_d(palette = "batlow") +
  theme_workshop(base_size = 16) +
  theme(legend.position = "none") +
  labs(
    y = "Life expectancy",
    color = "Continent",
    x = "Year"
  )

p_gdp <- ggplot(gap_continent, aes(year, mean_gdpPercap, color = continent)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = label_dollar(accuracy = 1)) +
  scale_color_scico_d(palette = "batlow") +
  theme_workshop(base_size = 16) +
  theme(legend.position = "none") +
  labs(
    y = "GDP per capita",
    color = "Continent",
    x = "Year"
  )
```

## Common multi-panel layouts

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 6.5
p_bubble_m4 /
  (p_gdp + p_life) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A", tag_suffix = ")")
```

## The patchwork package

:::{.nonincremental}

- Compose plots
- Collect shared legends and axes
- Tag panels
- Apply additional plot layers to every panel
- Place plots inside plots as insets

:::

The [patchwork documentation](https://patchwork.data-imaginist.com/) is excellent

## Live demo

Open `demo/03_patchwork.R`.

## Your turn {.inverse}

> :pencil2: **Exercise (~8 min):**
>
> Open `exercises/03_patchwork_exercise.R`


# 4. Exporting figures {.inverse}

```{r}
#| label: m5-setup
#| include: false
p_bubble_m5 <- p_bubble +
  scale_color_scico_d(palette = "batlow") +
  labs(color = "Continent", size = "Population")
```

## Same plot, different canvas

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 3.5
p_bubble_m5 + theme_workshop(base_size = 6)
```
:::

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 3.5
p_bubble_m5 + theme_workshop(base_size = 28)
```
:::

::::

## Rule of thumb

Starting points for Canvas and font sizes:

| Outlet | Canvas width | `base_size` |
|--------|--------------|-------------|
| Paper, single column | ~89 mm | 8–10 |
| Paper, double column | ~120–180 mm | 10–12 |
| Slide, half | ~150 mm | 14–16 |
| Slide, full | ~250 mm | 18–22 |
| Poster | depends on layout | 18–24 |

## Live demo

Open `demo/04_export.R`.

## Your turn {.inverse}

> :pencil2: **Exercise (~7 min):**
>
> Open `exercises/04_export_exercise.R`

# Other handy packages {.inverse}

## `gghighlight`

[`gghighlight`](https://yutannihilation.github.io/gghighlight/) to fade data and highlight specific elements.

```{r}
#| code-fold: true
library(gghighlight)
p_bubble +
  scale_color_scico_d(palette = "batlow") +
  theme_workshop() +
  gghighlight(
    continent %in% c("Europe", "Africa"),
    use_direct_label = FALSE
  )
```


## `ggrepel` + `ggtext`

[`ggrepel`](https://ggrepel.slowkow.com/) and [`ggtext`](https://wilkelab.org/ggtext/) for labels and markdown text

```{r}
#| code-fold: true
library(ggrepel)
library(ggtext)
ggplot(gap_continent, aes(year, mean_lifeExp, color = continent)) +
  geom_line(linewidth = 1) +
  geom_text_repel(
    data = gap_continent |> filter(year == max(year)),
    aes(label = continent),
    size = 6,
    nudge_x = 2,
    hjust = 0,
    direction = "y",
    seed = 1
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  scale_color_scico_d(palette = "batlow") +
  # use markdown in the title (see theme layer)
  labs(
    title = "*Life expectancy* by **continent**",
    x = "Year",
    y = "Life expectancy"
  ) +
  theme_workshop() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_markdown(),
    legend.position = "none"
  )
```


## `ggdist` raincloud plots

[`ggdist`](https://mjskay.github.io/ggdist/) to show distributions and uncertainty -> Barplot alternatives

```{r}
#| code-fold: true
library(ggdist)
ggplot(
  gap_2007 |> filter(continent != "Oceania"),
  aes(continent, gdpPercap, fill = continent)
) +
  stat_halfeye(
    adjust = 0.5,
    width = 0.6,
    .width = 0,
    justification = -0.3,
    point_colour = NA
  ) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5) +
  stat_dots(side = "left", justification = 1.1, dotsize = 0.4) +
  scale_y_log10(labels = label_dollar(accuracy = 1)) +
  scale_fill_scico_d(palette = "batlow") +
  theme_workshop() +
  theme(legend.position = "none")
```

## `ggridges`

[`ggridges`](https://wilkelab.org/ggridges/) for ridgeline distribution plots 

```{r}
#| code-fold: true
library(ggridges)
ggplot(gap_2007 |> filter(continent != "Oceania"), aes(lifeExp, continent, fill = continent)) +
  geom_density_ridges(alpha = 0.8, scale = 1.1) +
  scale_fill_scico_d(palette = "batlow") +
  theme_workshop() +
  theme(legend.position = "none") +
  labs(x = "Life expectancy (years)", y = NULL)
```

::: {.aside}

:::

## Takeaways

What you can take back to your own plots:

1. **Custom themes**: write one `theme_*()` function, source it, `theme_set()` it everywhere it's needed
2. **Colour with intent**: pick intuitive, colourblind-safe palettes (Okabe-Ito, viridis, scico), and check with `cvdPlot()`
3. **Multi-panel layouts**: compose plots with `patchwork`, collect shared legends, add tag panels, ...
4. **Export**: pick the canvas size, preview with `ggview::canvas()`, tweak `base_size` and geoms, and save with `ragg::ragg_png` or `cairo_pdf`.

## Next lecture/workshop

#### Topic tba

<br>

:date: 18.06.2026 :clock4: 4-5 p.m. :round_pushpin: Webex

:bell: [Subscribe to the mailing list](https://lists.fu-berlin.de/listinfo/toolsAndTips)

:e-mail: For topic suggestions and/or feedback [send me an email](mailto:selina.baldauf@fu-berlin.de)

# The end :) {background-color="#2c2c2c"}

Questions?
